Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-16 doc/invt.tex
add errscan.gp
invt.tex: edit to the first paragraph of Results

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL algorithm and metadynamics.
%
These methods update the bias potential
incrementally in regular intervals
and can easily achieve the target distribution.
%
However, they are less well understood theoretically,
as the updating loops
break the microscopic reversibility
of the underlying sampling mechanism.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods gradually
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
barducci2008}.
%
Thus, in late stages of a long simulation,
the updating is negligible and
the adaptive simulation tends to
the corresponding equilibrium one.
%
If the bias potential were converged to the exact value,
this particular equilibrium simulation can be thought as
the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal efficiency.

@@

For sufficiently small $\alpha(t)$,
the $v_i$'s remain roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_v}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \pi_i(t)
  =
  \frac{                p_i \, e^{-v_i(t)} }
       { \sum_{j = 1}^n p_j \, e^{-v_j(t)} }.
  \label{eq:h_ave}
\end{equation}
%
%
%
For the noise part, we have
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function of $t$
that vanishes at large $t$.
%
More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$

@@

Consider the following equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a generalized noise
that is invariant under time translation:
%
\begin{equation}
  \left\langle \xi(t) \, \xi(t') \right\rangle
  =
  \kappa(t - t')
  .
\notag
%\label{eq:noise_correlation}
\end{equation}
%
Additionally,
we assume that $\kappa(t)$ is an even function and
\begin{equation}
  \lim_{t \rightarrow \infty} \kappa(t)
  = 0
  .
  \label{eq:kappa_limit}
\end{equation}
%
For example, for a white noise,
$\kappa(t)$ is proportional to
Dirac's $\delta$-function, $\delta(t)$.
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }.
\label{eq:alpha_invt1}
\end{equation}

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(T)
=
x(0) \, e^{-q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\notag
%\label{eq:q_def}
\end{equation}
%
and
%
\begin{align}
u(q')
&\equiv
e^{q' - q(T)}.
\label{eq:u_def}
\end{align}
At the end of period $T$, we have
\begin{align}
&\left\langle x^2(T) \right\rangle
=
\langle x^2(0) \,
\rangle e^{-2 \, q(T)}
\notag
\\
&\qquad
+
\int_0^T \int_0^T
  \dot u\bigl( q(t) \bigr) \,
  \dot u\bigl( q(t') \bigr) \,
  \kappa(t - t') \, dt' \, dt.
\label{eq:x2t_average}
\end{align}

We now fix the endpoint value $q(T)$
and variate the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to variating $u$
with fixed endpoints,
$u\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$u\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the process.
%
%We shall further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
By the Euler-Lagrange equation, we get
$$
\begin{aligned}
0
&=
\frac{d}{dt} \int_0^T
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
\\
&= \int_0^T
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by assuming Eq. \eqref{eq:kappa_limit}
and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $t$,
we must have
%
\begin{equation}
  \ddot u\bigl( q(t') \bigr) = 0,
  \qquad
  \mathrm{or}
  \;\;
  \dot u\bigl( q(t') \bigr) = c,
\notag
%\label{eq:ddu_eq_0}
\end{equation}
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:u_def},
we get
$$
e^{ q(t') - q(T) }
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm,
and differentiating both sides
with respect to $t'$
yields Eq. \eqref{eq:alpha_invt1}.

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
  \label{eq:error_asym_Lagrangian}
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  ,
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
  %\label{eq:q_opt}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in th above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.

@@

\note{In practice, Eq. \eqref{eq:varY}
in inapplicable to zero eigenvalues, $\lambda_k$'s.
%
However, these eigenmodes can be ignored
as they do not contribute to the Lagrangian defined by
Eq. \eqref{eq:error_asym_Lagrangian}.
}

@@

Many commonly-used updating schemes
can be characterized by a rigid window
centered on the current bin.
%
We call these schemes homogeneous,
or translationally invariant.
%
An example is the Gaussian updating scheme
used in metadynamics.
%
Homogeneous updating schemes have significantly fewer
parameters than the general case,
and the multiplication of the updating matrix,
$\mathbf w$,
can be reduced to a convolution with the window,
or the updating kernel.
%
Thus,
the eigenvalues, $\lambda_k$'s, of $\mathbf w$
can be readily computed and manipulated
via the shape of the updating kernel.

@@

The above matrix, $\mathbf w$,
being symmetric to the indices $i$ and $j$,
satisfies Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance},
for the flat distribution.
%
To find the eigenvectors,
we define, for a periodic variable, $\phi_i$,
out-of-boundary values by
%
\begin{equation}
  \phi_i = \phi_{i \pm n},
\label{eq:phi_pbc}
\end{equation}
%
such that $i \pm n$ lies in between $1$ and $n$.
%
Then
%
\begin{equation}
  \sum_{ j = 1 }^n
    w_{ij} \, \phi_j
  =
  \sum_{ j = 1 - n }^{ 2 \, n }
    \mu_{i - j} \, \phi_j
  =
  \sum_{ l = -b }^{ b }
    \mu_l \, \phi_{ i - l}
  .
\label{eq:wmul_to_convol}
\end{equation}
%
\note{The derivation of the first step of
  Eq. \eqref{eq:wmul_to_convol}:
$$
\begin{aligned}
  \sum_{j = 1}^n
    w_{ij} \, \phi_j
  &=
  \sum_{j = 1}^n
    \mu_{i - j} \, \phi_j
  +
  \sum_{j = 1}^n
    \mu_{i - j - n} \, \phi_j
  +
  \sum_{j = 1}^n
    \mu_{i - j + n} \, \phi_j
  \\
  &=
  \sum_{j = 1}^n
    \mu_{i - j} \, \phi_j
  +
  \sum_{j = 1+n}^{2 \, n}
    \mu_{i - l} \, \phi_l
  +
  \sum_{j = 1-n}^0
    \mu_{i - j + n} \, \phi_j
  \\
  &=
  \sum_{j = 1-n}^{2 \, n}
    \mu_{i - j} \, \phi_j
  .
\end{aligned}
$$
The second step of Eq. \eqref{eq:wmul_to_convol}
follows from the constraint $\mu_l = 0$ for $|l| > b$.

@@

Similar to Eq. \eqref{eq:phi_pbc},
we define the out-of-boundary values
of $\phi_j$ as
%
\begin{equation}
  \phi_j
  =
  \begin{dcases}
    \phi_{ 1 - j }           & \mathrm{for \;} j \le 0, \\
    \phi_{ 2 \, n + 1 - j }  & \mathrm{for \;} j > n,
  \end{dcases}
\notag
%\label{eq:phi_refl}
\end{equation}
%
such that Eq. \eqref{eq:wmul_to_convol}
still holds.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \phi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \phi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \phi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \phi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \phi_j.
  \end{aligned}
  $$
}
%
Thus, we also have $n$ orthonormal eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
as
%
\begin{equation}
  \phi^{(k)}_i
  =
  \phi_{k i}
  =
  \frac{ \sqrt{ 2 - \delta_{k, 1} } }
       {             n              }
  \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, \pi}
            {                    n                           }
  ,
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
with eigenvalues
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \pi}{n}
  .
  \label{eq:wband_eigenvalue_refl}
\end{align}
%
\note{We derive the above two equations below.

  To simplify the calculation,
  let us first consider the unnormalized eigenvectors,
  $$
  \Phi^{(k)}_i
  =
  \cos \frac{ \left( i - \frac 1 2 \right) (k - 1) \, \pi}{n},
  $$
  which satisfies the reflective boundary condition,
  %Eq. \eqref{eq:phi_refl}.
  %
  So
  $$
  \begin{aligned}
  \sum_{i = 1}^n
    \Phi^{(k)}_i \, w_{ij}
  &=
  \sum_{l = -b}^b
    \Phi^{(k)}_{j - l} \, \mu_l
  \\
  &=
    \Phi^{(k)}_j \, \mu_0
  + \sum_{l=0}^{b}
    \mu_l \,
    \left[
      \Phi^{(k)}_{j-l}
      +
      \Phi^{(k)}_{j+l}
    \right]
  \\
  &= \Phi^{(k)}_j \, \lambda_k,
  \end{aligned}
  $$
  with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue_refl}.

@@

Note that Eq. \eqref{eq:wband_eigenvalue_refl}
differs from Eq. \eqref{eq:wband_eigenvalue_pbc}
by a factor of $2$
in the argument of the cosine function.
%
We may unify the two as
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2
  \sum_{ l = 1 }^b
    \mu_l \,
    \cos \frac{ (k - 1) \, l \, g \, \pi }
              {            n             }
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where the degeneracy $g$ is $2$
for the periodic case,
or $1$ otherwise.
%
Equation \eqref{eq:wband_eigenvalue}
shows that the eigenvalues, $\lambda_k$,
of the updating scheme
is the cosine transform of
the updating kernel.
%
So, for the updating scheme to be stable,
the cosine transform of the kernel
must be nonnegative.
%
If this is not so,
we can modify the kernel
using the technique described
in Sec. \ref{sec:stabilize_wband}.

@@

The Gaussian updating kernel is commonly
adopted by metadynamics simulations.
%
Below we show that the corresponding updating scheme
is stable, albeit less effective
in reducing short-wavelength noises.

We start by defining
a continuous limit
for $b = n - 1 \gg 1$:
$x = l \, \Delta x$,
and
$\mu_l = \mu(x) \, \Delta x$,
where
$\Delta x = \pi/n$
is the bin size.
%
Then
Eq. \eqref{eq:wband_eigenvalue}
can be approximated by an integral:
%
\begin{equation}
\lambda_k
=
2 \int_0^\pi
  \mu(x) \, \cos \bigl[ (k-1) \, x \bigr] \, dx,
\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
\begin{equation}
1 = 2 \int_0^\pi \mu(x) \, dx.
\label{eq:mx_normalization}
\end{equation}

>>>>

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL algorithm and metadynamics.
%
These methods update the bias potential
incrementally in regular intervals
and can easily achieve the target distribution.
%
However,
%they are less well understood theoretically, as
the updating loops break the microscopic reversibility
of the underlying sampling mechanism.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods gradually
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
barducci2008}.
%
Thus, in late stages of a long simulation,
the updating becomes negligible and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
if the bias potential were converged to the exact value,
this adaptive simulation tends to the ideal equilibrium simulation
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal efficiency.

@@

For sufficiently small $\alpha(t)$,
the $v_i$'s remain roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_v}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \pi_i(t)
  =
  \frac{                p_i \, e^{-v_i(t)} }
       { \sum_{j = 1}^n p_j \, e^{-v_j(t)} }.
  \label{eq:h_ave}
\end{equation}
%
%
%
For the noise part, we have
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function
that vanishes at large $t$.
%
\note{More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$
}%

@@

Consider the following equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a generalized noise
that is invariant under time translation:
%
$
  \left\langle \xi(t) \, \xi(t') \right\rangle
  =
  \kappa(t - t')
  .
$
%
Additionally,
we assume that $\kappa(t)$ is an even function and
\begin{equation}
  \lim_{t \rightarrow \infty} \kappa(t)
  = 0
  .
  \label{eq:kappa_limit}
\end{equation}
%
For example, for a white noise,
$\kappa(t)$ is proportional to
Dirac's $\delta$-function, $\delta(t)$.
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }.
\label{eq:alpha_invt1}
\end{equation}

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(T)
=
x(0) \, e^{-q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
u(q')
&\equiv
e^{q' - q(T)}.
\label{eq:u_def}
\end{align}
At the end of period $T$, we have
\begin{align}
&\left\langle x^2(T) \right\rangle
=
\langle x^2(0) \,
\rangle e^{-2 \, q(T)}
\notag
\\
&\qquad
+
\int_0^T \int_0^T
  \dot u\bigl( q(t) \bigr) \,
  \dot u\bigl( q(t') \bigr) \,
  \kappa(t - t') \, dt' \, dt.
\label{eq:x2t_average}
\end{align}

We now fix the endpoint value $q(T)$
and variate the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to variating $u$
with fixed endpoints,
$u\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$u\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the process.
%
%We shall further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
By the Euler-Lagrange equation, we get
$$
\begin{aligned}
0
&=
\frac{d}{dt} \int_0^T
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
\\
&= \int_0^T
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by assuming Eq. \eqref{eq:kappa_limit}
and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $t$,
we must have
%
$
\ddot u\bigl( q(t') \bigr) = 0
,
$
or
$
\dot u\bigl( q(t') \bigr) = c
,
$
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:u_def},
we get
$$
e^{ q(t') - q(T) }
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm,
and differentiating both sides
with respect to $t'$
yields Eq. \eqref{eq:alpha_invt1}.

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
%\label{eq:error_asym_Lagrangian}
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  ,
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
  %\label{eq:q_opt}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in th above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.

@@

\note{In practice, Eq. \eqref{eq:varY}
in inapplicable to zero eigenvalues, $\lambda_k$'s.
%
However, these eigenmodes can be ignored
as they do not contribute to the Lagrangian defined by
Eq. \eqref{eq:Lagrangian_const}.
}

@@

Many commonly-used updating schemes
can be characterized by a rigid window
centered on the current bin.
%
We call these schemes homogeneous,
or translationally invariant.
%
An example is the Gaussian updating scheme
used in metadynamics.
%
Homogeneous updating schemes have significantly fewer
parameters than the general case,
and the multiplication of the updating matrix,
$\mathbf w$,
can be reduced to a convolution with the window,
or the updating kernel.
%
Thus,
there is a one-to-one correspondence between
the eigenvalues of $\mathbf w$
and the updating kernel.

@@

The above matrix, $\mathbf w$,
being symmetric to the indices $i$ and $j$,
satisfies Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance},
for the flat distribution.
%
To find the eigenvectors,
we define, for a periodic variable, $\phi_i$,
the out-of-boundary values by
%
\begin{equation}
  \phi_i = \phi_{i \pm n},
\label{eq:phi_pbc}
\end{equation}
%
such that $i \pm n$ lies in between $1$ and $n$.
%
Then
%
\begin{equation}
  \sum_{ j = 1 }^n
    w_{ij} \, \phi_j
  =
  \sum_{ j = 1 - n }^{ 2 \, n }
    \mu_{i - j} \, \phi_j
  =
  \sum_{ l = -b }^{ b }
    \mu_l \, \phi_{ i - l}
  .
\label{eq:wmul_to_convol}
\end{equation}
%
\note{The derivation of the first step of
  Eq. \eqref{eq:wmul_to_convol}:
$$
\begin{aligned}
  \sum_{j = 1}^n
    w_{ij} \, \phi_j
  &=
  \sum_{j = 1}^n
    \mu_{i - j} \, \phi_j
  +
  \sum_{j = 1}^n
    \mu_{i - j - n} \, \phi_j
  +
  \sum_{j = 1}^n
    \mu_{i - j + n} \, \phi_j
  \\
  &=
  \sum_{j = 1}^n
    \mu_{i - j} \, \phi_j
  +
  \sum_{j = 1+n}^{2 \, n}
    \mu_{i - l} \, \phi_l
  +
  \sum_{j = 1-n}^0
    \mu_{i - j + n} \, \phi_j
  \\
  &=
  \sum_{j = 1-n}^{2 \, n}
    \mu_{i - j} \, \phi_j
  .
\end{aligned}
$$
The second step of Eq. \eqref{eq:wmul_to_convol}
follows from the constraint $\mu_l = 0$ for $|l| > b$.

@@

Parallel to Eq. \eqref{eq:phi_pbc},
we define the out-of-boundary values
of $\phi_i$ as
%
\begin{equation}
  \phi_i
  =
  \begin{dcases}
    \phi_{ 1 - i }           & \mathrm{for \;} i \le 0, \\
    \phi_{ 2 \, n + 1 - i }  & \mathrm{for \;} i > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
such that Eq. \eqref{eq:wmul_to_convol}
still holds.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \phi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \phi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \phi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \phi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \phi_j.
  \end{aligned}
  $$
}%
Thus, we also have $n$ orthonormal eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
compatible with Eq. \eqref{eq:phi_refl},
%
\begin{equation}
  \phi^{(k)}_i
  =
  \phi_{k i}
  =
  \frac{ \sqrt{ 2 - \delta_{k, 1} } }
       {             n              }
  \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, \pi}
            {                    n                           }
  ,
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
with eigenvalues
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \pi}{n}
  .
\label{eq:wband_eigenvalue_refl}
\end{align}
%
\note{Derivation.
  First consider the unnormalized eigenvectors,
  $$
  \Phi^{(k)}_i
  =
  \cos \frac{ \left( i - \frac 1 2 \right) (k - 1) \, \pi}{n},
  $$
  which satisfies the reflective boundary condition,
  Eq. \eqref{eq:phi_refl}.
  %
  So, by Eq. \eqref{eq:wmul_to_convol},
  $$
  \begin{aligned}
  \sum_{i = 1}^n
    \Phi^{(k)}_i \, w_{ij}
  &=
  \sum_{l = -b}^b
    \Phi^{(k)}_{j - l} \, \mu_l
  \\
  &=
    \Phi^{(k)}_j \, \mu_0
  + \sum_{l=0}^{b}
    \mu_l \,
    \left[
      \Phi^{(k)}_{j-l}
      +
      \Phi^{(k)}_{j+l}
    \right]
  \\
  &= \Phi^{(k)}_j \, \lambda_k,
  \end{aligned}
  $$
  with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue_refl}.

@@

Note that Eq. \eqref{eq:wband_eigenvalue_refl}
differs from Eq. \eqref{eq:wband_eigenvalue_pbc}
by a factor of $2$
in the argument of the cosine function.
%
We may unify the two as
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2
  \sum_{ l = 1 }^b
    \mu_l \,
    \cos \frac{ (k - 1) \, l \, g \, \pi }
              {            n             }
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where the degeneracy $g$ is $2$
for the periodic case,
or $1$ otherwise.
%
In both cases,
the eigenvalues, $\lambda_k$'s,
is the cosine transform of
the updating kernel.
%
For the updating scheme to be stable,
the cosine transform of the kernel
must be nonnegative.
%
\note{If this is not so,
we can modify the kernel
using the technique described
in Sec. \ref{sec:stabilize_wband}.}

@@

The Gaussian updating scheme is commonly
used in metadynamics simulations.
%
Below we shall show its stability
in the continuous limit.

For $n \gg 1$,
we define
$x = l \, g \, \Delta x$,
and
$\mu_l = \mu(x) \, \Delta x$,
where
$\Delta x = \pi/n$
is the bin size.
%
Then,
by assuming the largest $b \approx n/g$,
Eq. \eqref{eq:wband_eigenvalue}
can be approximated as an integral:
%
\begin{equation}
  \lambda_k
  =
  2 \int_0^\pi
    \mu(x) \, \cos \bigl[ (k-1) \, x \bigr] \, dx,
\notag
%\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
\begin{equation}
  1 = 2 \int_0^\pi \mu(x) \, dx.
\notag
%\label{eq:mx_normalization}
\end{equation}
